Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Chinook Salmon Populations

### 2022-02-02 analysis/paper/paper.Rmd
some discussion edits

All mark-recapture abundance estimators suggest a large population of Northern Pikeminnow occupy Deadwater Slough. The mean population size of Northern Pikeminnow in the Deadwater Slough was estimated to be ***value*** during the fall DSR emigration and ***value*** during the spring NRR emigration. Those estimates translate to densities (fish/km) of Northern Pikeminnow that are ***value*** and ***value*** times greater than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately ***value*** and ***value*** times greater than projections for the lower Columbia River [@Beamesderfer1996]. @Beamesderfer1991 acknowledge that their methods were unsuitable for sampling offshore, noting that water velocity, depth, irregular bottom contours, and barge traffic made sampling offshore ineffective; therefore it is possible that their estimates may have been low. Conversely, at Deadwater Slough the maximum depth was approximately 6 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Given the size (volume) of the Salmon River at the slough relative to the Columbia River, the estimated Northern Pikeminnow densities for Deadwater Slough are remarkable. Further, our estimate densities of Northern Pikeminnow suggest that slow water reaches outside of the reservoir complexes on the Snake and Columbia rivers may support exceptionally high predator densities, consistent with findings of @Harnish2014, @Gray2001, and @Zimmerman1999. Given the dearth of current data available in the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the Columbia River Basin; contemporary information on piscine predator abundances may be needed [@Widener2021].

<!--
The mean population size of Northern Pikeminnow in the Deadwater Slough during the fall season was estimated to be `r prettyNum(unique(round(impact_df$N[impact_df$season == "Fall"])), big.mark = ",")` and during the spring 2021 to be `r prettyNum(unique(round(impact_df$N[impact_df$season == "Spring"])), big.mark = ",")`. Those estimates translate to linear densities of `r prettyNum(round(unique(impact_df$N[impact_df$season == "Fall"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(unique(impact_df$N[impact_df$season == "Spring"]) / ddw_rkms), big.mark = ",")` 
which is `r round(unique(impact_df$N[impact_df$season == "Fall"]) / ddw_rkms / john_day_dens)` and `r round(unique(impact_df$N[impact_df$season == "Spring"]) / ddw_rkms / john_day_dens)` times greater, respectively, than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r as.character(english(round(unique(impact_df$N[impact_df$season == "Fall"]) / ddw_rkms / low_columbia_fish_km)))` and `r as.character(english(round(unique(impact_df$N[impact_df$season == "Spring"]) / ddw_rkms / low_columbia_fish_km)))` times greater than projections for the lower Columbia River [@Beamesderfer1996]. Although estimated densities for Deadwater Slough are high given the relative size (volume) of the Salmon River at the slough relative to the Columbia River, high CPUE combined with low recapture rates (one in every `r round(nrow(cmr_df) / sum(cmr_df$capture_type == "Recapture", na.rm = T))` captured fish were recaptures) suggest this area may support exceptionally high predator densities. Further, @Beamesderfer1991 acknowledge that their methods were unsuitable for sampling offshore, noting that water velocity, depth, irregular bottom contours, and barge traffic made sampling offshore ineffective. Conversely, at Deadwater Slough the maximum depth was approximately 10-15 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Our estimates of Northern Pikeminnow areal density are also higher than those reported elsewhere in the Columbia River basin. We estimate the density of Northern Pikeminnow in Deadwater Slough, as fish per hectare, to be `r prettyNum(round(unique(impact_df$N[impact_df$season == "Fall"]) / ddw_hectares), big.mark = ",")` for fall and `r prettyNum(round(unique(impact_df$N[impact_df$season == "Spring"]) / ddw_hectares), big.mark = ",")` for spring. For comparison, the only other available estimates of areal density, that the authors are aware of, are 4.4 fish per hectare for John Day Reservoir [@Beamesderfer1991] and 15 fish per hectare in Lake Washington [@Bartoo1977]. Given the dearth of current data available in the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the Columbia River Basin. @Widener2021 has identified a need for more current information on piscine predator populations in the Columbia River.
-->

Several assumptions in our estimators may have influenced the magnitude of our abundance results. First, all models assumed a closed population, however, it is possible that immigration and emigration occurred. Nevertheless, our estimators still provide unbiased estimates of abundance assuming the immigration and emigration rates are equal between marked and unmarked fish. Given the size of the sampling area, and the short duration of our surveys, that assumption is likely met. The assumption of a closed population can also affect results if mortality occurs for some marked fish released back to the population. During our field processing, no mortalities or injury post release was observed; however, the potential for mortality amongst marked and released fish cannot be discounted. Next, our estimates assumed that capture and recapture events are random samples of the population. Angling methods may have a size selection bias, thereby limiting our abundance estimates to a size range susceptible to angling, resulting in a conservative estimate of the total population. We also question the validity of our assumption of equal catchability of individual fish between sampling events. During our study, we recaptured two individual fish during multiple surveys indicating that marked fish were still susceptible to angling.<!--We are also comfortable with the assumptions of no loss of marks or no mistaken or missed marks given that marks were physical (hole punches in fins) and that each fish was examined on a measuring board to collect lengths.--> However, we suspect that marked and released fish may not have mixed thoroughly back into the population prior to recapture events. Although we attempted to release marked fish evenly throughout Deadwater Slough, we anecdotally observed anglers congregating in particular areas of the slough to socialize or exploit the "good" fishing locales. Violations of the assumption of equal catchability between marked and unmarked fish may lead to an overestimate of abundance of Northern Pikeminnow in our study if marked fish have lower catchability. This bias will be proportional to the difference in catchability of marked and unmarked fish, so if marked fish are only 70% as likely to be caught as unmarked fish (due to behavior or spatial distribution) the true abundance is closer to 70% of our abundance estimate.

<!--
All mark-recapture abundance estimators suggest a large number of Northern Pikeminnow occupying Deadwater Slough. Estimators assumed: (1) the population is closed (no immigration, emigration, births, or deaths during the sampling period), (2) all fish have equal chance of being caught in the second (and subsequent) samples(s), (3) marking a fish does not affect its chances of recapture, (4) no loss of marks, and (5) marks are not missed or mistaken. If the population is actually open (potential immigration and/or emigration during the survey period), our estimators should still provide unbiased estimates of abundance as long as the immigration and emigration rates are equal between marked and unmarked fish. Given the size of the sampling area, and the limited length of the sampling period (two weeks), we are comfortable with the assumption of no immigration or emigration. Assumption (3) indicates whether the second (and subsequent) surveys are random samples of the population. Angling methods may provide a size selection bias, but that only affects the interpretation of the abundance estimates (i.e., our estimates are the abundance of Northern Pikeminnow susceptible to angling above some size threshold). Unequal catchability between individual fish is very difficult to assess, especially if that heterogeneity arises from being caught once; however, we did recapture two individual fish more than once suggesting that marked fish were still susceptible to angling. We are also comfortable with assumptions 4 and 5 (no loss of marks and no mistaken or missed marks) given that marks were physical (hole punches in fins) and that each fish was examined on a measuring board to collect lengths.

We do consider two potential violations of assumption, both which reduces the probability that a marked fish is susceptible to capture in the second (and subsequent) samples. First, the assumption of a closed population could be violated if some marked fish that are released back to the population die. We don't believe this case to be likely, as fish that were dead or nearly so after processing were not released back to the population (nor recorded as such), and those that were released were hand-released only if they were able to swim away on their own volition; however, the potential for marked and released fish to have died at a higher rate than unmarked fish should be acknowledged. The second, and more likely, reason is that marked and released fish did not mix thoroughly back into the population (at least not at a fast enough rate), reducing their probability of capture in subsequent captures, thus violating assumption (2). We attempted to release marked fish throughout Deadwater Slough to facilitate mixing back into the population; however, anecdotally volunteer anglers would congregate to particular areas of the slough either to socialize with other anglers, because the fishing was good, etc. and if marked and released fish were not readily available in those areas that would make marked fish less susceptible to capture. This violation would, in effect, reduce the number of marked fish available for recapture or reduce the marking fraction in some areas of the slough. Violations of assumption (2) would lead to an overestimate of abundance of Northern Pikeminnow which is a real possibility in our study.
-->

@@

In addition to the large population, the observed PSD (***value***%) in this study was large compared to observations of 41% below Bonneville Dam and 18% in Bonneville Reservoir [@Winther2020]. This suggests that a larger fraction of Northern Pikeminnow in Deadwater Slough are of a quality size relative to populations reported elsewhere in the Columbia River. Notably, the state catch-and-release record Northern Pikeminnow, measuring ***value*** mm TL, was caught in Deadwater Slough during the fall 2020 survey.

@@

Recovery of Chinook Salmon in the Upper Salmon MPG is desired to support local fisheries and economies. At least five of the eight populations must meet criteria set forth by @McElhany2000 and the @ICTRT2007 for the MPG to be considered viable and for recovery of the Snake River Evolutionary Significant Unit. We estimate that ***value*** juvenile Chinook Salmon will be consumed by Northern Pikeminnow, which translates to ***value*** "adult equivalents" which is approximately ***value***% of contemporary returns to the MPG. Considering the diet scenarios used in our study are likely conservative assumptions during the fall DSR emigration, the estimated number of Chinook Salmon consumed is substantial.

Spawner abundance is perhaps the most important metric considered in determining a population's viability. We estimated that consumption of juvenile Chinook Salmon by Northern Pikeminnow in the Deadwater Slough potentially reduces natural-origin returns to upriver populations by ***value*** to ***value*** adults, annually. This is between ***value***% and ***value***% of the total mean adult escapement<!--of 558--> reported for the Upper Salmon MPG in 2017 - 2019 [@Kinzer2020]. In addition to juvenile Chinook Salmon, juvenile steelhead and Sockeye Salmon, including hatchery releases, are also likely prey items for Northern Pikeminnow in Deadwater Slough. Accordingly, we surmise that adult returns of steelhead and Sockeye Salmon are affected by predation, including for hatchery populations which provide recreational fishing opportunities. Consequently, reducing predation mortality at Deadwater Slough could potentially benefit multiple upriver natural and hatchery populations, including ESA-listed species. Because the Deadwater Slough is part of the migratory pathway for multiple species and populations of emigrating salmonids, the positive impact could be greater than individual tributary rehabilitation actions which typically benefit a single population.

@@

<!--
What about adding cover? It could potentially provide refugia for juvenile salmonids, but could make recreational angling more difficult
-->

>>>>

All mark-recapture abundance estimators suggest a large population of Northern Pikeminnow occupy Deadwater Slough. The mean population size of Northern Pikeminnow was estimated to be `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` during the fall DSR emigration and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` during the spring NRR emigration. Those estimates translate to linear densities (fish/rkm) that are `r round((mean(all_est$N[all_est$model == "Schnabel"]) / ddw_rkms) / john_day_dens)` and `r round((spring_abund$N_S / ddw_rkms) / john_day_dens)` times greater than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r round((mean(all_est$N[all_est$model == "Schnabel"]) / ddw_rkms) / low_columbia_fish_km)` and `r round((spring_abund$N_S / ddw_rkms) / low_columbia_fish_km)` times greater than projections for the lower Columbia River [@Beamesderfer1996]. Our areal densities (fish/100 $m^2$) for Deadwater Slough are `r round((mean(all_est$N[all_est$model == "Schnabel"]) / ddw_m2) * 100, 1)` and `r round((spring_abund$N_S / ddw_m2) * 100, 1)` for fall and spring, respectively. @Beamesderfer1991 acknowledge that their methods were unsuitable for sampling offshore, noting that water velocity, depth, irregular bottom contours, and barge traffic made sampling offshore ineffective; therefore it is possible that their estimates may have been low. Conversely, at Deadwater Slough the maximum depth was approximately 6 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Given the size (volume) of the Salmon River at the slough relative to the Columbia River, the estimated Northern Pikeminnow densities for Deadwater Slough are remarkable. Further, our estimate densities of Northern Pikeminnow suggest that slow water reaches outside of the reservoir complexes on the Snake and Columbia rivers may support exceptionally high predator densities, consistent with findings of @Harnish2014, @Gray2001, and @Zimmerman1999. Given the dearth of current data available in the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the Columbia River Basin; contemporary information on piscine predator abundances may be needed [@Widener2021].

Several assumptions in our estimators may have influenced the magnitude of our abundance results. First, all models assumed a closed population, however, it is possible that immigration and emigration occurred. Nevertheless, our estimators still provide unbiased estimates of abundance assuming the immigration and emigration rates are equal between marked and unmarked fish. Given the size of the sampling area, and the short duration of our surveys, that assumption is likely met. We also caught one individual that ws confirmed tagged in Deadwater Slough in a previous year suggesting that at least some individuals remain in the reach or at least migrate to Deadwater Slough multiple times, annually. The assumption of a closed population can also affect results if mortality occurs for some marked fish released back to the population. During field processing, no mortalities or injury post release was observed; however, the potential for mortality amongst marked and released fish cannot be discounted. Next, our estimates assumed that capture and recapture events are random samples of the population. Angling methods may have a size selection bias, thereby limiting our abundance estimates to a size range susceptible to angling, resulting in a conservative estimate of the total population. We also question the validity of our assumption of equal catchability of individual fish between sampling events. During our study, we recaptured two individual fish during multiple days within a survey indicating that marked fish were still susceptible to angling. We are also comfortable with the assumptions of no loss of marks or no mistaken or missed marks given that marks were physical (hole punches in fins) and that each fish was examined on a measuring board to collect lengths. However, we suspect that marked and released fish may not have mixed thoroughly back into the population prior to recapture events. Although we attempted to release marked fish evenly throughout Deadwater Slough, we anecdotally observed anglers congregating in particular areas of the slough to socialize or exploit the "good" fishing locales. Violations of the assumption of equal catchability between marked and unmarked fish may lead to an overestimate of abundance of Northern Pikeminnow in our study if marked fish have lower catchability. This bias will be proportional to the difference in catchability of marked and unmarked fish. As an example, if marked fish are only 70% as likely to be caught as unmarked fish (due to behavior or spatial distribution) the true abundance is closer to 70% of our abundance estimate.

@@

In addition to the large population, the observed PSD (`r round(psd)`%) in this study was large compared to observations of 41% below Bonneville Dam and 18% in Bonneville Reservoir [@Winther2020]. This suggests that a larger fraction of Northern Pikeminnow in Deadwater Slough are of a quality size relative to populations reported elsewhere in the Columbia River. Notably, the state catch-and-release record Northern Pikeminnow, measuring `r max(gastric_df$length, na.rm = T)` mm TL, was caught in Deadwater Slough during the fall 2020 survey.

@@

Recovery of Chinook Salmon in the Upper Salmon MPG is desired to support local fisheries and economies. At least five of the eight populations must meet criteria set forth by @McElhany2000 and the @ICTRT2007 for the MPG to be considered viable and for recovery of the Snake River Evolutionary Significant Unit. We estimate that `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` juvenile Chinook Salmon may be consumed by Northern Pikeminnow assuming that 60% of their diet is fish and 50% of that is Chinook Salmon. Considering that that diet scenarios used in our study are likely conservative assumptions, especially during the fall DSR emigration, the estimated number of Chinook Salmon consumed is substantial.

Spawner abundance is perhaps the most important metric considered in determining a population's viability. We estimated that consumption of juvenile Chinook Salmon by Northern Pikeminnow in the Deadwater Slough potentially reduces natural-origin returns to upriver populations by `r round(total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50])` (95% CI: `r round(total_impacts$adult_Lci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50])` - `r round(total_impacts$adult_Uci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50])`) adults, annually. This is `r round((total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50] / 558) * 100)`% of the total mean adult escapement estimated<!--of 558--> for the Upper Salmon MPG in 2017 - 2019 [@Kinzer2020]. Even under the most conservative scenario where 30% of the Northern Pikeminnow diet is fish and 30% of that is Chinook Salmon we estimate the "adult equivalents" of the juveniles consumed to be `r round(total_impacts$adult_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30])` (95% CI: `r round(total_impacts$adult_Lci[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30])` - `r round(total_impacts$adult_Uci[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30])`) which is `r round((total_impacts$adult_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30] / 558) * 100)`% or recent natural-origin adult escapements. In addition to juvenile Chinook Salmon, juvenile steelhead and Sockeye Salmon, including hatchery releases, are also likely prey items for Northern Pikeminnow in Deadwater Slough. Accordingly, we surmise that adult returns of steelhead and Sockeye Salmon are affected by predation, including for hatchery populations which provide recreational fishing opportunities. Consequently, reducing predation mortality at Deadwater Slough could potentially benefit multiple upriver natural and hatchery populations, including ESA-listed species. Because the Deadwater Slough is part of the migratory pathway for multiple species and populations of emigrating salmonids, the positive impact could be greater than individual tributary rehabilitation actions which typically benefit a single population.

@@
